Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-08-11 manu_draft.Rmd
added trend method comparison data object and text results

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(chgtrnd)
data(mods_chl18)
data(mods_chl24)
data(mods_chl27)
data(mods_chl32)
data(mods_chl34)

@@

Nitrogen and phospohorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.  However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.  Historical resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.  Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations beginning in 1999 (Figure \@ref(fig:obsdat)) [@Cloern20]. Although visual changes in observed data are apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed, particularly on a seasonal basis.

The analysis evaluated near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].  Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an intercept term $s_0$.  The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.  Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.  All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods [@Wood17].

@@

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  A comparison of GAM performance between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation [@Box64].  The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.  The profile vector for each chlorophyll time series was obtained using the *boxcox* function from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.  Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:

@@

where $g()$ is the transformation function for the appropriate GAM at the esitmated $\lambda$ value. Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.  The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlorophyll time series with the optimal $\lambda$ value.  Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.

All models were compared using standard summary statistics describing overall model fit to the observed data.  In addition to the Generalized Cross-Validatoin score provided by the mgcv package, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.  Comparisons were made between the different GAMs and for the transformations used to define the $g()$ function for chlorophyll at each station. Model predictions were based on the standard model output from the *predict()* function from the mgcv package.  A long-term trend independent of seasonal variation was also estimated for each model by subtracting the seasonal terms (i.e., $s_1(doy)$ and $ti_1(doy, year)$) from the predictions. Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used. For hypothesis tests that required estimates of central tendency and confidence intervals for mean values, back-transformations were based on an exponentiation or inverse function of the transformed response variable as before, with the addition of a correction factor.  This bias-correction factor was simply the variance of the model residuals divided by two for variables in log-space and correction with a forecast variance estimate as specified in the *InvBoxCox()* function from the forecast package.

@@

Although the seasonal averages obtained from GAMs described the expected values for each year, describing the direction and magnitude of the seasonal change across years was also of interest.  Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected seasonal value of chlorophyll and the predictor is year as a numeric value.  Although this approach could be used to estimate a seasonal rate of change, the linear fit through the seasonal means does not account for the variance in the seasonal averages.  To address this issue, we use mixed effects regression meta-analysis to evaluate seasonal changes and their uncertainty over a finite set of years.  This analysis is an extension of conventional regression by accounting for uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:

@@

where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and deviation of the seasonal mean from the true mean ($S_i$).  Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.  The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].  Results have a similar interpretation as those from regression analysis (i.e., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.

@@

Each of the three GAM structures (constant season, interactive season, and constant season*) were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.  For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoother terms.  Models were also evaluated using different transformations to determine which had improved fit for specific time series.  Trend analyses with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.  Temporal and spatial patterns within and between stations were also idenfied to better quanity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.

@@

A total of six models (three GAM structures, two link functions) were fit to each of nine stations, producing summary statistics for 18 different models (Table \@ref(tab:modprftab)).  Chlorophyll trends were generally well-explained across all models and stations, with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (`r minstmod` `r minsttrn` at station `r minst`) to `r maxr2`% (`r maxstmod` `r maxsttrn` at station `r maxst`). The constant season model had the worst performance for all stations (average R$^2$ `r avecsr2`%), with either the interactive season or constant season* model having improved measures of model fit.  For all but stations 18 and 22, the constant season* model provided significantly lower AIC and GCV scores and higher R-squared values compared to the interactive season models.  For station 18, the interactive season and constant season* models provided only a marginal improvement over the constant season model.  For station 22, the interactive season and constant season* models provided similar but much improved fit to the data as compared to the constant season model.  Moel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance. Station 18 was the only location where the three GAM models were not significantly different from each other for the Box-Cox transformation.

GAM predictions from north to south on the longitidunal axis of South Bay provided different descriptions of annual and seasonal changes in chlorophyll.  Four representative stations are shown in Figure \@ref(fig:trnddat) that capture spatial variation in chlorophyll and different temporal trends described by the models.  All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record. Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location.  However, the magnitude of the fall peak at more southern stations shifted over time depending on the model.  In particular, the interactive season model indicates a shift in the magnitude of the fall peak at stations 24, 27, and 32 with maximum concentrations observed around the mid-early 2000s. The annual trends for all models suggested similar trends over time, i.e., an increase followed by a recent decrease, whereas the constant season* models with an unconstrained smoother on the `decYear` predictor varied considerably at station 24, 27, and 32. Annual trends at these stations were difficult to generalize and had substantial overlap with the seasonal trends.  Conversely, station 18 had minimal visual differences in annual and seasonal trends across all models, which is reflected by similarities in the model performance statistics in Table \@ref(tab:modprftab).

@@

Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to July and August to December (Figure \@ref(fig:trnddat)).  This figure illustrates the narrative decriptions that can be obtained for each test and the differences that can be observed depending on season.  Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.  However, the change from 2000 to 2010 was statistically significant with an increase of 78%.  The period from 2010 to 2017 showed a significant decrease of 39.8%.  An assessment of the annual differences indicaates if changes overall were observe, but provides no information on seasonal shifts that may have contributed to these differences. The results in figure \@ref(fig:trnddat))d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences. Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.  Similarly, rows g-i show the seasonal averages and trend estimates from August to December.  The seasonal trend results suggest that increase in annual averages from 2000 to 2010 was likely causd by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).  Seasonal changes from 2010 to 2017 showed that neither the January to July or August to December changes had significant changes, suggesting an overall decrease in chlorophyll regardless of season may have recently occurred.

Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).  From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.  The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significiant.  From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).  From 2010 to 2016, all four southern stations had significant decreases in annual chlorophyll.  The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).  For the January to July period, significant changes were only observed at the southern stations with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.  No significant changes were observed for the August to December period.

A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).  Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).  However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.  Trends at staiton 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.  Station 34 had signifificant increases in the JFM period from 1990-2010 and all but the OND (October, November, December) period for 2000 to 2010.  Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.  Station 36 had significant chlorophyll increases in all seasons except OND for the 1990 to 2000 period.  Accordingly, table \@ref(tab:seastrndtab) demosntrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the alternative seasonal blocks shown in Figure \@ref(fig:trndmap)b.

@@

@@

```{r sumtab, tab.cap = 'Station locations, sample sizes, and median values for salnity, dissolved inorganic nitrogen (DIN, sum of ammonia, nitrate, and nitrite), and chlorophyll.  Rows are arranged from north to south.'}
totab <- rawdat %>% 
  filter(param %in% c('sal', 'din', 'chl')) %>% 
  group_by(station, param) %>% 
  summarise(
    n = n(), 
    medv = median(value, na.rm = T)
  ) %>% 
  ungroup() %>% 
  gather('var', 'val', n, medv) %>% 
  unite('var', param, var) %>% 
  mutate(var = factor(var, levels = c('sal_n', 'sal_medv', 'din_n', 'din_medv', 'chl_n', 'chl_medv'))) %>% 
  left_join(locs, by = 'station') %>% 
  spread(var, val)

@@

```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using BoxCox and logarathmic transformations. Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `Const seas`),  $\\Delta$ GCV, and R-squared values.  Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`Const seas` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  ns: not significant at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}

>>>>

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(chgtrnd)
data(cmptrnd)
data(mods_chl18)
data(mods_chl24)
data(mods_chl27)
data(mods_chl32)
data(mods_chl34)

@@

Nitrogen and phosphorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.  However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.  Historical resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.  Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations beginning in 1999 (Figure \@ref(fig:obsdat)) [@Cloern20]. Although visual changes in observed data are apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed, particularly on a seasonal basis.

The analysis evaluated near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].  Discrete chlorophyll concentrations at each station were determined by fluorimetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an intercept term $s_0$.  The smoothing functions are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.  Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.  All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods [@Wood17].

@@

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  A comparison of GAM performance between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisfying assumptions for GAMs.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation [@Box64].  The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.  The profile vector for each chlorophyll time series was obtained using the *boxcox* function from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.  Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:

@@

where $g()$ is the transformation function for the appropriate GAM at the estimated $\lambda$ value. Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.  The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlorophyll time series with the optimal $\lambda$ value.  Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.

All models were compared using standard summary statistics describing overall model fit to the observed data.  In addition to the Generalized Cross-Validation score provided by the mgcv package, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.  Comparisons were made between the different GAMs and for the transformations used to define the $g()$ function for chlorophyll at each station. Model predictions were based on the standard model output from the *predict()* function from the mgcv package.  A long-term trend independent of seasonal variation was also estimated for each model by subtracting the seasonal terms (i.e., $s_1(doy)$ and $ti_1(doy, year)$) from the predictions. Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used. For hypothesis tests that required estimates of central tendency and confidence intervals for mean values, back-transformations were based on an exponentiation or inverse function of the transformed response variable as before, with the addition of a correction factor.  This bias-correction factor was simply the variance of the model residuals divided by two for variables in log-space and correction with a forecast variance estimate as specified in the *InvBoxCox()* function from the forecast package.

@@

Although the seasonal averages obtained from GAMs described the expected values for each year, describing the direction and magnitude of the seasonal change across years was also of interest.  Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected seasonal value of chlorophyll and the predictor is year as a numeric value.  Although this approach could be used to estimate a seasonal rate of change, the linear fit through the seasonal means does not account for the variance in the seasonal averages.  To address this issue, we use mixed effects regression meta-analysis (mixed-meta hereafter) to evaluate seasonal changes and their uncertainty over a finite set of years.  This analysis is an extension of conventional regression by accounting for uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:

@@

where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and deviation of the seasonal mean from the true mean ($S_i$).  Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chlorophyll averages.  The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].  Results have a similar interpretation as those from regression analysis (i.e., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.

@@

Each of the three GAM structures (constant season, interactive season, and constant season*) were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.  For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoother terms.  Models were also evaluated using different transformations to determine which had improved fit for specific time series.  Trend analyses with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.  Temporal and spatial patterns within and between stations were also identified to better quantity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.

@@

A total of six models (three GAM structures, two link functions) were fit to each of nine stations, producing summary statistics for 18 different models (Table \@ref(tab:modprftab)).  Chlorophyll trends were generally well-explained across all models and stations, with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (`r minstmod` `r minsttrn` at station `r minst`) to `r maxr2`% (`r maxstmod` `r maxsttrn` at station `r maxst`). The constant season model had the worst performance for all stations (average R$^2$ `r avecsr2`%), with either the interactive season or constant season* model having improved measures of model fit.  For all but stations 18 and 22, the constant season* model provided significantly lower AIC and GCV scores and higher R-squared values compared to the interactive season models.  For station 18, the interactive season and constant season* models provided only a marginal improvement over the constant season model.  For station 22, the interactive season and constant season* models provided similar but much improved fit to the data as compared to the constant season model.  Moeel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance. Station 18 was the only location where the three GAM models were not significantly different from each other for the Box-Cox transformation.

GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll.  Four representative stations are shown in Figure \@ref(fig:trnddat) that capture spatial variation in chlorophyll and different temporal trends described by the models.  All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record. Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location.  However, the magnitude of the fall peak at more southern stations shifted over time depending on the model.  In particular, the interactive season model indicates a shift in the magnitude of the fall peak at stations 24, 27, and 32 with maximum concentrations observed around the mid-early 2000s. The annual trends for all models suggested similar trends over time, i.e., an increase followed by a recent decrease, whereas the constant season* models with an unconstrained smoother on the `decYear` predictor varied considerably at station 24, 27, and 32. Annual trends at these stations were difficult to generalize and had substantial overlap with the seasonal trends.  Conversely, station 18 had minimal visual differences in annual and seasonal trends across all models, which is reflected by similarities in the model performance statistics in Table \@ref(tab:modprftab).

## Trend estimates

Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to July and August to December (Figure \@ref(fig:trnddat)).  This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.  Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.  However, the change from 2000 to 2010 was statistically significant with an increase of 78%.  The period from 2010 to 2017 showed a significant decrease of 39.8%.  An assessment of the annual differences indicated if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences. The results in figure \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences. Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.  Similarly, rows g-i show the seasonal averages and trend estimates from August to December.  The seasonal trend results suggested that the increase in annual averages from 2000 to 2010 was likely caused by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).  Seasonal changes from 2010 to 2017 showed that neither the January to July nor August to December changes were significant, suggesting an overall decrease in chlorophyll across all seasons may have recently occurred.

Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).  From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.  The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.  From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).  From 2010 to 2016, all four southern stations had significant decreases in annual chlorophyll.  The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).  For the January to July period, significant changes were only observed at the southern stations, with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.  No significant changes were observed for the August to December period.

A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).  Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).  However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.  Trends at station 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.  Station 34 had significant increases in the JFM period from 1990-2010 and all other periods the OND (October, November, December) period for 2000 to 2010.  Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.  Station 36 had significant chlorophyll increases in all seasons, except OND for the 1990 to 2000 period.  Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.

@@

Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.  Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.  In addition to mixed-meta analysis, two alternative trend methods were used where the first shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a linear regression model on annual averages derived from the GAM seasonal estimates.  The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.

The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).  The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.  Slope estimates from the linear model applied to the averages from the GAMs as compared ot the mixed-meta results were identical at almost all stations, excluding station 32 in the January-July period from 1990-2000 when the former showed a negative slope and the latter showed a positive slope.  The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.  Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.

@@

```{r, results = 'hide', message = F}
toplo <- cmptrnd
p <- ggplot(toplo, aes(y = station, x = yrcoef, shape = mod, colour = pval)) + 
  facet_grid(seas ~ yrs, scales = 'free_x') + 
  geom_point(size = 3, position = position_dodge(width = 0.5), alpha = 0.6) + 
  scale_shape('Trend model') + 
  scale_colour_discrete_diverging(name = 'Trend', palette = 'BlueRed2') +
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  theme_minimal() + 
  theme(
    strip.background = element_blank(),
    panel.background = element_rect(fill = 'gray98', colour = NA), 
    panel.border = element_blank(), 
    legend.position = 'top'
    ) +
  labs(
    y = 'Station (north to south)', 
    x = expression(paste("Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
  )

png('figs/trndcmp.png', height = 5, width = 7.5, family = 'serif', units = 'in', res = 400)
p
dev.off()
```
```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at each station. The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM constant season* model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM constant season* model.  Values for each model are the slope esimates as annual change per year within each season, with color denoting significant trends.'}
knitr::include_graphics('figs/trndcmp.png')
```

@@

```{r sumtab, tab.cap = 'Station locations, sample sizes, and median values for salinity, dissolved inorganic nitrogen (DIN, sum of ammonia, nitrate, and nitrite), and chlorophyll.  Rows are arranged from north to south.'}
totab <- rawdat %>% 
  filter(param %in% c('sal', 'din', 'chl')) %>% 
  group_by(station, param) %>% 
  summarise(
    n = n(), 
    medv = median(value, na.rm = T)
  ) %>% 
  ungroup() %>% 
  gather('var', 'val', n, medv) %>% 
  unite('var', param, var) %>% 
  mutate(var = factor(var, levels = c('sal_n', 'sal_medv', 'din_n', 'din_medv', 'chl_n', 'chl_medv'))) %>% 
  left_join(locs, by = 'station') %>% 
  spread(var, val)

@@

```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)). Results are for models of chlorophyll-a using BoxCox and logarithmic transformations. Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `Const seas`),  $\\Delta$ GCV, and R-squared values.  Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.  The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`Const seas` is the base model and does not have a p-value).  A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.  ns: not significant at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
